---
title: "Introduction to statswift"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to statswift}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

The **statswift** package is a lightweight, easy-to-use toolkit for descriptive statistics and data summaries in R. It is designed for students and analysts who need comprehensive statistical output quickly, without relying on heavy external packages.

`statswift` provides five core functions:

| Function | Purpose |
|---|---|
| `swift_summary()` | Extended descriptive statistics for a numeric vector |
| `swift_outliers()` | Outlier detection using IQR or Z-score methods |
| `swift_compare()` | Group-wise descriptive comparison |
| `swift_distplot()` | Distribution visualization (histogram + density + boxplot) |
| `swift_report()` | Full statistical report for an entire data frame |

It uses **S3 classes** (`SwiftSummary`, `SwiftOutliers`, `SwiftCompare`) and one **S4 class** (`SwiftReport`), each with clean print/show methods.

---

## Installation

```{r install, eval=FALSE}
# Install from GitHub using devtools
devtools::install_github("yourusername/statswift")
```

```{r load}
library(statswift)
```

---

## 1. Extended Descriptive Summary — `swift_summary()`

The `swift_summary()` function computes far more than base R's `summary()`. It includes skewness, kurtosis, standard error, coefficient of variation, and more.

```{r swift_summary}
x <- c(23, 45, 67, 12, 89, 34, 56, NA, 78, 90, 11, 45, 67)
result <- swift_summary(x)
print(result)
```

The output is an **S3 object** of class `SwiftSummary`. You can access individual statistics directly:

```{r swift_summary_access}
result$mean
result$skewness
result$cv
```

**Interpreting skewness:**

- Near 0: roughly symmetric
- Positive: right-skewed (tail to the right)
- Negative: left-skewed (tail to the left)

---

## 2. Outlier Detection — `swift_outliers()`

`swift_outliers()` supports two classic methods:

- **IQR method** (default): flags values below Q1 − 1.5×IQR or above Q3 + 1.5×IQR
- **Z-score method**: flags values with |Z| > threshold (default 3)

```{r swift_outliers_iqr}
y <- c(2, 3, 4, 5, 6, 7, 100, -50, 5, 4, 3, 6)
swift_outliers(y)
```

```{r swift_outliers_zscore}
swift_outliers(y, method = "zscore", threshold = 2.5)
```

The returned `SwiftOutliers` object tells you the outlier values, their positions in the vector, and the percentage of data flagged.

---

## 3. Group Comparison — `swift_compare()`

Use `swift_compare()` to compare a numeric variable across categorical groups. This is especially useful before running group-based tests (e.g., t-tests, ANOVA).

```{r swift_compare}
scores <- c(85, 90, 78, 92, 88, 76, 95, 60, 70, 82, 91, 74)
groups <- c("A","A","B","A","B","B","A","B","B","A","A","B")

swift_compare(scores, groups)
```

The result is a clean data frame showing N, Mean, Median, SD, Min, Max, and IQR for each group.

---

## 4. Distribution Plot — `swift_distplot()`

`swift_distplot()` creates a two-panel visualization: a histogram with overlaid density curve on top, and a boxplot below.

```{r swift_distplot, fig.width=6, fig.height=5}
set.seed(123)
data <- c(rnorm(80, mean = 50, sd = 10), rnorm(20, mean = 80, sd = 5))
swift_distplot(data, main = "Bimodal Distribution Example", col = "steelblue")
```

Vertical lines for the **mean** (red dashed) and **median** (green dotted) are shown by default — their separation is a visual cue for skewness.

---

## 5. Full Dataset Report — `swift_report()` (S4)

For a holistic view of an entire data frame, use `swift_report()`. It applies descriptive statistics to every numeric column and returns a structured **S4 object** of class `SwiftReport`.

```{r swift_report}
report <- swift_report(mtcars, data_name = "Motor Trend Cars (1974)")
show(report)
```

You can also inspect the S4 slots directly:

```{r swift_report_slots}
report@n_rows
report@numeric_cols
report@summaries[["mpg"]]
```

---

## Putting It All Together

Here is a typical workflow using `statswift` on the built-in `airquality` dataset:

```{r workflow}
data(airquality)

# Step 1: Get an overall report
swift_report(airquality, data_name = "New York Air Quality (1973)")

# Step 2: Dive into Ozone specifically
swift_summary(airquality$Ozone)

# Step 3: Check for outliers
swift_outliers(airquality$Ozone, method = "iqr")

# Step 4: Compare Ozone by Month
swift_compare(airquality$Ozone, airquality$Month)
```

```{r workflow_plot, fig.width=6, fig.height=5}
# Step 5: Visualize the distribution
swift_distplot(airquality$Ozone, main = "Ozone Distribution (NYC, 1973)", col = "darkorange")
```

---

## Class System Summary

| Class | System | Constructor | Print/Show |
|---|---|---|---|
| `SwiftSummary` | S3 | `swift_summary()` | `print.SwiftSummary()` |
| `SwiftOutliers` | S3 | `swift_outliers()` | `print.SwiftOutliers()` |
| `SwiftCompare` | S3 | `swift_compare()` | `print.SwiftCompare()` |
| `SwiftReport` | S4 | `swift_report()` | `show("SwiftReport")` |

---

## License

`statswift` is licensed under the **MIT License**. See the `LICENSE` file for details.
